---
title: "Validation"
author: "Tina Lasisi"
date: "`r Sys.time()`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}

library(tidyverse)
library(cowplot)
library(knitr)
library(truncnorm)
library(reshape2)
library(data.table)

knitr::opts_chunk$set(echo = FALSE, include = TRUE, eval = TRUE, warning = FALSE, message = FALSE, fig.retina = 2, fig.width = 8, fig.height = 4)

F = rprojroot::is_rstudio_project$make_fix_file()

```

```{r functions, include=FALSE}
plot_path = F("output")

pltsave_func <- function(plot, plot_path, width, height){
  filename <- paste0(deparse(substitute(plot)), ".png")
  fullpath <- file.path(plot_path, filename)
  ggsave(
      filename = filename,
      plot = plot, 
      path = plot_path,
      width = width,
      height = height)
  plot(plot)
}
```

# Curvature 
Here, we will evaluate the accuracy of *fibermorph* in estimating the length and curvature of hair using simulated data. See simulation script [here](https://github.com/tinalasisi/2020_HairPheno_manuscript/blob/main/code/simhair.R). 

The simulated data can be found [here](https://github.com/tinalasisi/2020_HairPheno_manuscript/blob/main/data/concat_simcurvature_Nov152020.csv). 


```{r df_curvature, include=FALSE}

# Read simulated data and format it for plotting.
# 
# I added IDs for the images and gave consistent naming to simulated vs. estimated values. I also added an element ID (elem_ID) referring to the ID of each hair within an image

dat <- fread(F("data/fibermorph_out_simcurvature_11212020_concat.txt"))
colnames(dat) = c("filename", "null","curvature","length")

dat$filename = str_remove_all(dat$filename, 
                              ".csv")

dat <- dat%>%
  mutate(ID = filename)%>%
  separate(filename, 
           into = c("ImageSum",
                    "ID_sim",
                    "sim_radius"),
           sep = "_") %>%
  select(-c(ImageSum,ID_sim,null))%>%
  mutate(sim_length = pi/2,
         sim_radius = as.numeric(sim_radius),
         sim_curvature = 1/sim_radius,
         est_radius = 1/curvature) %>%
  select(ID, everything()) %>% 
  rename(est_curvature = curvature,
         est_length = length) %>%
  group_by(ID) %>% 
  mutate(elem_ID = as.character(row_number())) %>% 
  pivot_longer(cols = -c(ID, elem_ID), 
               names_to = c("sim", "var"), 
               names_pattern = "(.*)_(.*)") %>%
  select(ID, elem_ID, var, sim, value) %>% 
  pivot_wider(names_from = sim, values_from = value)

```  

We simulated arcs of various curvatures at a length of 1.57mm. There were 25 arcs per image. 

## Simulated vs. estimated curvature & length
To calculate the accuracy of our measurements, we compared the known parameters with the parameters estimated from our fibermorph package.

```{r plt_curvature_corr_length}
# Plot estimated curvature vs true curvature

plt_curvature <- dat %>% 
  filter(var == "curvature") %>% 
  ggplot(aes(sim,est))+
  geom_abline(intercept = 0, 
              slope = 1, 
              color = "red")+
    stat_smooth(method = "lm")+
  geom_point(color = "black", 
             alpha = 0.5, 
             size = 0.8)+
  scale_color_viridis_c()+
  theme_classic()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = bquote("Estimated curvature ("*mm^-1*")"),
       title = "Estimated vs simulated hair curvature")


# Plot the distribution of estimated length of each hair. 
# 
plt_length_histo <- dat %>% 
  filter(var == "length") %>% 
  ggplot(aes(est))+
  geom_histogram(color = "black",
                 fill = "grey")+
  geom_vline(xintercept = pi/2, 
             color = "red", linetype = "dashed")+
  theme_classic()+
  labs(y = "Count",
       x = bquote("Estimated length ("*mm*")"),
       title = "Distribution estimated hair length")


plt_curv_length_corr_hist = plot_grid(plt_curvature, plt_length_histo, labels = c("a", "b"))


plt_curv_length_corr_hist

```

In Fig.a we see that there is a near perfect correlation between the simulated and estimated curvatures. Fig.b shows the distribution of estimated hair lengths around the simulated length (red line). 

We plot simulated curvature against estimated length to show the distribution of estimated length as a function of curvature.

```{r plt_length_curv, fig.height=6, fig.width=7}

plt_length_curv <- dat %>% 
  pivot_wider(id_cols = c(ID, elem_ID), names_from = var, values_from = c(sim, est))%>% 
  ggplot(aes(sim_curvature, est_length))+
    stat_smooth(method = "lm")+
  geom_hline(yintercept = pi/2, 
             color = "red", linetype = "dashed") +
  geom_point(color = "black", 
             alpha = 0.5, 
             size = 0.8)+
  scale_color_viridis_c()+
  theme_classic()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = bquote("Estimated length in mm"),
       title = "Simulated curvature vs estimated length")

plt_length_curv
```

There seems to be a broader range of error in the estimation of length in straighter hairs. This is likely a result of the majority of pixels being oriented in a manner that causes a divergence between the pixel length (number of pixels) and the real length that is being measured. We apply a correction for this known issue in image analysis, however, it is expected that there will still be some error. Note that each point in this figure represents an individual hair fragment within an image. This supports the notion that it is not the low curvature per se, but rather the combination of low curvature and specific orientations that increases the error in length estimation.

## Measurement error in curvature and length

In addition to the correlations between simulated and estimated parameters, we calculate root mean square error (RMSE) and percent error as alternatives to investigate the measurement error of our package.

NB: we present the data summarized for each image (i.e. all 25 fragments) as we cannot provide a hair fragment to hair fragment comparison.

```{r df_curv_error, include=FALSE}

dat.se <- dat %>%
  group_by(ID, sim, var) %>%
  summarize(hair_count = max(as.numeric(elem_ID)),
         error = mean(est - sim),
         abs.error = mean(abs(error)),
         rmse = sqrt(sum(error^2))/hair_count,
         percent.error = abs.error/mean(sim)*100,
         hair_count = n()) %>%
  distinct()

```

### Error statistics

Below, we calculate the mean error values for both RMSE and percent error. 

```{r stat_error_curv, echo=FALSE, results='asis'}

err_var <- dat.se %>%
  group_by(var) %>% 
  summarize(mean.rmse = mean(rmse),
            perent.error= mean(percent.error))

kable(err_var, caption = "RMSE and Percent Error per variable")

```
We see less than 1% error across the variables and RMSE of less than 0.0005.

Below, we plot the data.

### Root mean square error
First, we plot the root mean square error for curvature and length. 

```{r plt_rmse_curv}

plt_rmse_curvature <- dat.se %>% 
  filter(var == "curvature") %>% 
  ggplot(., 
       aes(sim, rmse))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = "RMSE",
       title = "Error in Curvature estimation")

length_rmse_mean = mean(dat.se$rmse[dat.se$var=="length"])

plt_rmse_length <- dat.se %>% 
  filter(var == "length") %>% 
  ggplot(., 
       aes(rmse))+
  geom_histogram(bins=10,
                 color = "black",
                 fill = "grey")+
  geom_vline(xintercept = length_rmse_mean, 
             color = "blue")+
  theme_classic()+
  labs(y = "Count",
       x = "RMSE",
       title = "Error in length estimation")


plt_rmse = plot_grid(plt_rmse_curvature, plt_rmse_length, labels = c("a", "b"))

plt_rmse

```

We then examine the relationship between curvature and RMSE of length

```{r plt_length_curv_rmse, fig.height=5, fig.width=7}

plt_length_curv_rmse <- dat.se %>% 
  pivot_wider(id_cols = c(ID), names_from = var, values_from = c(sim, rmse, rmse))%>% 
  ggplot(aes(sim_curvature, rmse_length))+
    stat_smooth(method = "lm")+
  geom_point(color = "black", 
             alpha = 0.5, 
             size = 0.8)+
  scale_color_viridis_c()+
  theme_classic()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = bquote("RMSE in length"),
       title = "Simulated curvature vs RMSE estimated length")

plt_length_curv_rmse

```

We observe an increase in RMSE with curvature. 

### Percent error

Below we plot the percent error for curvature and length. 

```{r plt_perror_curv}

plt_perror_curvature <- dat.se %>% 
  filter(var == "curvature") %>% 
  ggplot(., 
       aes(sim, percent.error))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = "Percent error",
       title = "Error in Curvature estimation")

length_perror_mean = mean(dat.se$percent.error[dat.se$var=="length"])

plt_perror_length <- dat.se %>% 
  filter(var == "length") %>% 
  ggplot(., 
       aes(percent.error))+
  geom_histogram(bins=10,
                 color = "black",
                 fill = "grey")+
  geom_vline(xintercept = length_perror_mean, 
             color = "blue")+
  theme_classic()+
  labs(y = "Count",
       x = "Percent error",
       title = "Error in length estimation")


plt_perror = plot_grid(plt_perror_curvature, plt_perror_length, labels = c("a", "b"))

plt_perror

```


```{r plt_length_curv_perror, fig.height=5, fig.width=7}

plt_length_curv_perror <- dat.se %>% 
  pivot_wider(id_cols = c(ID), names_from = var, values_from = c(sim, rmse, percent.error))%>% 
  ggplot(aes(sim_curvature, percent.error_length))+
    stat_smooth(method = "lm")+
  geom_point(color = "black", 
             alpha = 0.5, 
             size = 0.8)+
  scale_color_viridis_c()+
  theme_classic()+
  labs(x = bquote("Simulated curvature ("*mm^-1*")"),
       y = bquote("Percent error in length"),
       title = "Simulated curvature vs percent error estimated length")

plt_length_curv_perror

```
Here we see that error appears to increase slightly with curvature if considering the data in terms of percent error.


# Cross-section

The fibermorph section function estimates area, minimum diameter, maximum diameter and eccentricity for a given cross-sectional image. We tested the measurement error using randomly generated circles and non-circular ellipses.

```{r df_eccentricity_simulations, include=FALSE}

df_section <- read_csv(F("data/df_section_validation.csv"))

```

## Correlation between simulated and estimated section parameters


```{r plt_section_correlation, fig.height=8, fig.width=8}
plt_area = ggplot(df_section, aes(ref_area,area))+
  geom_point()+
  theme_classic()+
  labs(x = "Area (simulated)",
       y = "Area (estimated)",
       title = "Estimated vs simulated area")+
  geom_abline(intercept = 0, 
              slope = 1, color="red")

plt_ecc = ggplot(df_section, aes(ref_eccentricity,
                          eccentricity))+
  geom_point()+
  theme_classic()+
  labs(x = "Eccentricity (simulated)",
       y = "Eccentricity (estimated)",
       title = "Estimated vs simulated eccentricity")+
  geom_abline(intercept = 0, slope = 1, color="red")


plt_min = ggplot(df_section, aes(ref_min_diam,min))+
  geom_point()+
  theme_classic()+
  labs(x = "Minimum diameter (simulated)",
       y = "Minimum diameter (estimated)",
       title = "Estimated vs \nsimulated minimum diameter")+
  geom_abline(intercept = 0, 
              slope = 1, color="red")

plt_max = ggplot(df_section, aes(ref_max_diam,
                          max))+
  geom_point()+
  theme_classic()+
  labs(x = "Maximum diameter (simulated)",
       y = "Maximum (estimated)",
       title = "Estimated vs \nsimulated maximum diameter")+
  geom_abline(intercept = 0, slope = 1, color="red")


plt_sect_corr <- plot_grid(plt_area, plt_ecc, plt_min, plt_max, labels = c("a", "b", "c", "d"))

plt_sect_corr
```

We see strong correlations between the estimated and simulated values for each cross-sectional parameter. 

```{r imsave_val, include=FALSE}
plt_curv_sect_corr <- plot_grid(plt_curvature, plt_length_histo, plt_area, plt_ecc, labels = c("a", "b", "c", "d"))


pltsave_func(plot = plt_curv_sect_corr, plot_path = plot_path, width = 8, height = 8)

```

## Measurement error for cross-sectional parameters
We calculate the percent error and RMSE for the cross-sectional parameters.

```{r df_section_error, include=FALSE}

# Calculated error for each image comparing the reference value with the estimated value
# 
section_se <- df_section %>% 
  pivot_longer(cols = -ID) %>% 
  separate(col = name, into = c("ref", "var"), sep = "_", fill = "left") %>% 
  mutate(ref = replace_na(ref, "est")) %>% 
  pivot_wider(names_from = ref, values_from = value) %>% 
  mutate(error = ref - est,
         abs_error = abs(error),
         mean_error = mean(error),
         mean_abs_error = mean(abs_error),
         rmse = sqrt((ref - est)^2),
         percent_error = (abs(ref - est) / est) *100,
         var = as.factor(var))


```

First, we calculate mean error values for all parameters.

```{r stat_error_section, results='asis'}

err_var <- section_se %>%
  group_by(var) %>% 
  summarize(mean_rmse = mean(rmse),
            perent_error= mean(percent_error, na.rm = TRUE))

kable(err_var, caption = "RMSE and Percent Error per variable")
```

Percent error is considerably under 0.02% for each of the parameters with RMSE under 0.01 for all but area.

As one of the simulated ellipses was a circle with an eccentricity of 0, any deviation from this produces an infinite percent error. So below we present the values removing this observation.

```{r stat_error_section_filt, results='asis'}

err_var <- section_se %>%
  group_by(var) %>% 
  filter(percent_error != Inf) %>% 
  summarize(mean_rmse = mean(rmse),
            perent_error= mean(percent_error, na.rm = TRUE))

kable(err_var, caption = "RMSE and Percent Error per variable")
```

### Root mean square error
Below, we plot RMSE as a function of each parameter.

```{r plt_section_RMSE, fig.height=8, fig.width=8}


plt_rmse_area <- section_se %>% 
  filter(var == "area") %>% 
  ggplot(., 
       aes(ref, rmse))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated area ("*mm^-1*")"),
       y = "RMSE",
       title = "RMSE in area estimation")


plt_rmse_eccentricity <- section_se %>% 
  filter(var == "eccentricity") %>% 
  ggplot(., 
       aes(ref, rmse))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated eccentricity ("*mm^-1*")"),
       y = "RMSE",
       title = "RMSE in eccentricity estimation")

plt_rmse_min <- section_se %>% 
  filter(var == "min") %>% 
  ggplot(., 
       aes(ref, rmse))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated min ("*mm^-1*")"),
       y = "RMSE",
       title = "RMSE in min estimation")


plt_rmse_max <- section_se %>% 
  filter(var == "max") %>% 
  ggplot(., 
       aes(ref, rmse))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated max ("*mm^-1*")"),
       y = "RMSE",
       title = "RMSE in max estimation")


plt_rmse_section <- plot_grid(plt_rmse_area, plt_rmse_eccentricity, plt_rmse_min, plt_rmse_max, labels = c("a", "b", "c", "d"))

plt_rmse_section


```

There does not appear to be any overarching pattern in RMSE across the variables.

### Percent error
Below we plot the correlation between simulated values and percent error for each parameter. 

```{r plt_section_corr, fig.height=8, fig.width=8}

plt_perror_area <- section_se %>% 
  filter(var == "area") %>% 
  ggplot(., 
       aes(ref, percent_error))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("simulated area ("*mm^-1*")"),
       y = "Percent error",
       title = "Error in area estimation")


plt_perror_eccentricity <- section_se %>% 
  filter(var == "eccentricity") %>% 
  ggplot(., 
       aes(ref, percent_error))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated eccentricity ("*mm^-1*")"),
       y = "Percent error",
       title = "Error in eccentricity estimation")

plt_perror_min <- section_se %>% 
  filter(var == "min") %>% 
  ggplot(., 
       aes(ref, percent_error))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated min ("*mm^-1*")"),
       y = "Percent error",
       title = "Error in min estimation")

plt_perror_max <- section_se %>% 
  filter(var == "max") %>% 
  ggplot(., 
       aes(ref, percent_error))+
  geom_point()+
  theme_classic()+
  stat_smooth()+
  labs(x = bquote("Simulated max ("*mm^-1*")"),
       y = "Percent error",
       title = "Error in max estimation")


plt_error_section <- plot_grid(plt_perror_area, plt_perror_eccentricity, plt_perror_min, plt_perror_max, labels = c("a", "b", "c", "d"))

plt_error_section

```

We observe a general decrease in percent error for each parameter. 



